---
title: "Lab 7 Assignment: Group 27"
format:
  html:
    embed-resources: true
editor: visual
---

+-----------------------+-----------------------+---------------+---------------+
| Mikkel Piester Hansen | Christian Christensen | Dilan\        | Dea\          |
|                       |                       | Cifci         | Bajo          |
+-----------------------+-----------------------+---------------+---------------+
| s223105               | s223096               | s233597       | s232105       |
+-----------------------+-----------------------+---------------+---------------+

# Load library

```{r}
library("tidyverse")
library(broom)
library(cowplot)

```

# Load data

```{r}
raw_dir <- "data/_raw/"
data_file <- "gravier.RData"
data_loc <- "https://github.com/ramhiser/datamicroarray/raw/master/data/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}
load(file = str_c(raw_dir, data_file))
```

# 1. Clean data

Binding cols of the gravier dataset

```{r}
gravier_clean <- gravier |>
  bind_cols() |>
  as_tibble()
```

As the dataset is very large and contains 2900+ genes, we would like to just test the PCA analysis with the first 10 genes in the data.

```{r}
gravier_subset <- gravier_clean |> 
  select(1:10, last_col())
```

# 2. Data in PC coordinates

Fitting PCA to numeric coloumns in the data

```{r}
pca_fit <- 
  gravier_subset |> 
  select(where(is.numeric)) |> 
  prcomp(scale = TRUE)
```

Plotting PC coordinates

```{r}
pca_fit |> 
  augment(gravier_subset) |> 
  ggplot(aes(.fittedPC1, 
             .fittedPC2, 
             color = y)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(good = "#D55E00", 
               poor = "#0072B2")
  ) +
  theme_half_open(12) + 
  background_grid()
```

# 2. PC coordinates rotation matrix 

Extract rotation matrix

```{r}
pca_fit |> 
  tidy(matrix = "rotation")
```

Plotting the rotation matrix

```{r}

arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)


pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() +
  theme_minimal_grid(12)
```

# 3. Looking at variance explained by PC components

```{r}
pca_fit |> 
  tidy(matrix = "eigenvalues")
```

Plotting the variance

```{r}
pca_fit |> 
  tidy(matrix = "eigenvalues") |> 
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
```
